---
title: "Reverse Engineering Project"
author: "Devon Milley, Timothy Dashiell, Emmett Gartner"
date: "03/12/2022"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this notebook, we are reverse engineering the story, L.A. is slammed with record costs for legal payouts  (https://www.latimes.com/local/lanow/la-me-ln-city-payouts-20180627-story.html)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
install.packages("tidyverse")
options(scipen=999)
install.packages("janitor")
library(tidyverse)
library(janitor)

```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data.

```{r}
# Load required data

la_payouts <- read.csv("data/la_city_legal_payouts.csv")
View(la_payouts)

# Clean required data and prepare for analysis if needed.

la_payouts <- la_payouts %>% mutate(AMOUNT=as.numeric(parse_number(AMOUNT)))



```

## Sentences to Engineer
In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1

* **Sentence text**: “Last budget year, the total for all types of payouts reached $201 million — the highest on record — but city officials are unsure whether that is the largest amount because records go back only two decades.”

* **General steps to reverse engineer**:
1. Downloaded an excel sheet of all LA City Payouts between 2006-2018 from LA Times Github, deleted extraneous titles that could get in the way when converted to .csv.
2. Converted the file to a .csv.
3. Removed the /$ from the amount field.
4. Converted the amount field to a numeric datatype.
5. Grouped cases by year, summarized total payout amounts for each year.
6. Arranged the results by the greatest total.


* **Analysis summary**: The results of our analysis confirm what the LA Times reported. According to the results of our reverse engineering, Los Angeles' total for all types of payouts was more than $201 million, which is the highest it's ever been. [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here
#sum payouts for 2018

la_payouts%>%
  group_by(Year)%>%
    summarize(
    count = n(),
    total = sum(AMOUNT)) %>%
  arrange(desc(total))


# Display results of code below this codeblock

```

### Sentence 2

* **Sentence text**: “Thousands of legal battles involving the Los Angeles Police Department, including a rash of recent settlements over police shootings and wrongful imprisonment, have added up to more than 40% of the total — a higher sum than any other city department, the analysis showed. That includes more than $24 million paid to two men who were freed after decades in prison, along with lawsuits over arrests, injuries, car crashes and alleged retaliation and discrimination against its own employees.”

* **General steps to reverse engineer**:
1. To determine that the L.A. Police Department has accounted for 40% of the total L.A. cash settlements, we must:
  a. Determine the total settlement amounts for every department.
  b. Determine the total settlement amount for all of L.A.
2. We start by grouping the settlement amounts by 'DEPARTMENT' and organizing them in descending order.
3. We then determine the total settlement amount by summarizing all amounts greater than 0.
4. Lastly, we create a new column that accounts for percentage share of each department (through dividing the department's aggregate by the total settlement amount)

* **Analysis summary**: Our results show thousands of legal cases involving the police department have added up to 42.26% of LA's total payout costs, which matches what the LA Times found. We also found that [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here
la_payouts%>%
  filter(AMOUNT > 0)%>%
  summarize(
    total= sum(AMOUNT))

la_payouts_departments<-la_payouts%>%
  group_by(DEPARTMENT)%>%
    summarize(
    case_count = n(),
    total_amount = sum(AMOUNT)) %>%
  arrange(desc(total_amount))

<<<<<<< HEAD
la_payouts%>%
  filter(AMOUNT > 0) %>% 
  summarize(
    total_amount_of_all_cases = sum(AMOUNT))
=======
la_payouts_pct_share <- la_payouts_departments%>%
  mutate(pct_share =(total/886979072))
>>>>>>> 1e3f19eefb425d64eaa44ecb70bcfb181f86d957

  

<<<<<<< HEAD
# Display results of code below this codeblock
(374824384/886979072)*100

la_payouts %>% 
  filter(DEPARTMENT=="Police Department") %>% 
  arrange(desc(AMOUNT))
=======
>>>>>>> 1e3f19eefb425d64eaa44ecb70bcfb181f86d957

# Display results of code below this codeblock
la_payouts_pct_share
```

### Sentence 3

* **Sentence text**: “Lawsuits tied to the city departments that oversee its streets and sidewalks, including the Bureau of Street Services and the Department of Transportation, accounted for a growing share. Legal payouts for “dangerous conditions” have rivaled and, last budget year, even exceeded the cost of lawsuits over police misconduct, the analysis found. In May, the city attorney’s office warned that there had been “unprecedented growth” in such payouts.”


* **General steps to reverse engineer**:
1. For this sentence, we must determine whether L.A. payouts for "dangerous conditions" accounted for more costs than the L.A. Police Department in 2018.
2. To do so, we filter all cases by year (2018) and with the Case.Type description of "dangerous conditions."
3. Next, we summarize the total amount of settlements that meet those conditions.
4. Lastly, we do the same process for 2018 L.A. Police Department cases and compare the totals of each data frame.

* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here
la_payouts%>%
  filter(Year == 2018, (str_detect(CASE.TYPE, 'Dangerous Condition')))%>%
    summarize(
    count = n(),
    total = sum(AMOUNT)) %>%
  arrange(desc(total))

la_payouts%>%
  filter(Year == 2018 , DEPARTMENT == 'Police Department')%>%
    summarize(
    count = n(),
    total = sum(AMOUNT))%>%
        arrange(desc(total))

# Display results of code below this codeblock

```

### Sentence 4

* **Sentence text**: Many cases have arisen from people being wounded or killed while crossing city streets: In recent years, the city agreed to pay 15 dollars million to end a legal battle over a man who was gravely injured while crossing Franklin Avenue in Hollywood, another 15 million dollars to the family of two girls hit in a North Hollywood crosswalk and $9.5 million to the family of a teenage girl fatally struck by a taxi near Dockweiler State Beach.


* **General steps to reverse engineer**:


* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here
la_payouts_traffic <- la_payouts%>%
  filter(DEPARTMENT == 'Department of Transportation' | (str_detect(CASE.TYPE, 'Traffic')) | DEPARTMENT == 'DEPARTMENT OF TRANSPORTATION')%>%
  arrange(desc(AMOUNT))

# Display results of code below this codeblock

```

### Sentence 5

* **Sentence text**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

-30-
