---
title: "Data Analysis Project"
author: "Devon Milley, Timothy Dashiell, Emmett Gartner"
date: "05/10/2022"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction/Summary

In this notebook, we are analyzing The Lasting Legacy Of Redlining.

## Variable explanations
HOLC Grades
The Home Owners Loan Corporation was a New Deal-era federal agency that collaborated with local banks, cities and real estate appraisers to rate the “quality” of neighborhoods on a scale of A-D, with A being the best and D the worst. Factors that went into determining a neighborhood’s grade included objective indicators such as access to amenities like parks and transportation, and biased variables based on the economic class and race of residents.

HOLC grades were then used to determine a homeowner’s qualifications for a mortgage; homeowners in D-grade neighborhoods were less likely to receive friendly mortgages and reduced the value of their homes. The long-term impacts of HOLC grades included a drain of wealth from neighborhoods with unfavorable grades.

HOLC grades will be used as a backdrop for the other variables in our data set. We will focus on how a neighborhood's HOLC grade impacted its current’ racial breakdown.


Demographic Data
Accompanying the HOLC grades in FiveThirtyEight’s data set are modern demographic data that show the racial makeup of HOLC neighborhoods and surrounding areas.

zone-block-matches.csv
The other .csv file in this data set is the product of FiveThirtyEight journalists matching old HOLC grades/neighborhoods with modern census blocks, then deriving census data and populating census geometry with it.


## Load libraries
```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(sf)
library(janitor)
library(lubridate)
library(dplyr)
library(tidycensus)
library(tigris)
```

## Load and Cleaning Data

There are 551 rows and 28 columns in the data set.
```{r}
# Load required data

metro_grades<-read_csv("data/metro-grades.csv")

zone_block_matches<-st_read("data/zone-block-matches.csv")

```
## Mean racial composition percentages for A, B, C, and D HOLC Grades:
We started analyzing our data broadly, looking to see the racial composition of each HOLC grade. To do this, we grouped the metro_grades data set by HOLC grade and found the mean of each race’s percentage. 

We found that the majority of neighborhoods that received a HOLC grade of A had a high percent of white people. A-grade neighborhoods only had about 8% Black and Hispanic populations each. Lower graded neighborhoods at the C and D level, had higher percents of Black and Hispanic populations, making up about 51% combined for D-level neighborhoods and 41% combined for C-level neighborhoods.  
```{r}
holc_grades_race <- metro_grades %>% 
  group_by(holc_grade) %>% 
  summarise(
    pct_white = mean(pct_white),
    pct_black = mean(pct_black),
    pct_hisp = mean(pct_hisp),
    pct_asian = mean(pct_asian),
    pct_other = mean(pct_other))
```

## Mean racial composition percentages for A, B, C, and D HOLC Grades by Region:
When analyzing racial breakdown by region, as defined by the U.S. Census, we found the West North Central region has an average white population percentage of HOLC D grades of 58.2%, which is significantly higher than the other regions. The states included in this region are Iowa, Kansas, Minnesota, Missouri, Nebraska, North Dakota and South Dakota.

The average white population percentage of HOLC D grades in this region can be explained by population density and the area's racial breakdown. According to Statista, all states in the West North Central region rank in the bottom half of states' population density. Each state in that region also has an average white population percent of at least 61%. 

We also found that the East South Central region, which includes Kentucky, Tennessee, Mississippi and Alabama, had the highest average percent of white population with A-level HOLC grades at 81.28%. That is 8% higher than the overall average white population percentage of HOLC A grades.

Half of the states in this region have an average white population percent of more than 50%. The other two states — Alabama and Mississippi — have 49% and 25% average white population percent, respectively. It will be interesting to look at HOLC grades in Alabama and Mississippi to see whether there is a higher percent of A-grade neighborhoods with a high average Black population percent because the states themselves have a higher Black population in general.

The West South Central region has the highest average Hispanic population percent of HOLC A grades at 21.4%. This is about 18% higher than the overall average. States in this region include Texas, Oklahoma, Arkansas and Louisiana. The high Hispanic population in Texas — 46% — probably brought up the region's average. 

The West South Central region has the highest average Black population percent of HOLC A grades at 12.6%. That is 4% higher than the overall average Black population percentage of HOLC A grades. 

To draw these conclusions, we created a new region column using regions defined by the U.S. Census (https://www2.census.gov/geo/pdfs/maps-data/maps/reference/us_regdiv.pdf). We then found the average race of each HOLC grade by region. To compare our results to the racial breakdown on a regional level and state level, we created different data frames for each region that averaged out the percent of each race. 

```{r}
metro_grades_plus_state <- metro_grades %>% 
  mutate(
    state=str_sub(metro_area, start=-2, end=-1)
  )
```

```{r}
metro_grades_west_north_central <- metro_grades_plus_state %>% 
  filter(state == "IA" | state== "KS" | state=="MN" | state=="MO" | state=="NE" | state=="ND" | state=="SD") %>% 
  group_by(state) %>% 
  summarise(
    avg_white_pct = mean(pct_white),
    avg_black_pct = mean(pct_black),
    avg_hisp_pct = mean(pct_hisp),
    avg_asian_pct = mean(pct_asian),
    avg_other_pct = mean(pct_other)
  )
```

```{r}
metro_grades_east_south_central <- metro_grades_plus_state %>% 
  filter(state == "KY" | state== "TN" | state=="MS" | state=="AL") %>% 
  group_by(state) %>% 
  summarise(
    avg_white_pct = mean(pct_white),
    avg_black_pct = mean(pct_black),
    avg_hisp_pct = mean(pct_hisp),
    avg_asian_pct = mean(pct_asian),
    avg_other_pct = mean(pct_other)
  )
```

```{r}
metro_grades_west_south_central <- metro_grades_plus_state %>% 
  filter(state == "TX" | state== "OK" | state=="AR" | state=="LA") %>% 
  group_by(state) %>% 
  summarise(
    avg_white_pct = mean(pct_white),
    avg_black_pct = mean(pct_black),
    avg_hisp_pct = mean(pct_hisp),
    avg_asian_pct = mean(pct_asian),
    avg_other_pct = mean(pct_other)
  )
```


```{r}

metro_grades_w_region <- metro_grades_plus_state %>% 
  mutate(
    region = case_when(
      (state %in% c("CA", "OR", "WA", "AK", "HI")) == TRUE ~ "Pacific West",
      (state %in% c("ID", "MT", "WY", "NV", "UT", "CO", "AZ", "NM")) == TRUE ~ "Mountain West",
      (state %in% c("TX", "OK", "AR", "LA")) == TRUE ~ "West South Central",
      (state %in% c("ND", "SD", "NE", "KS", "MN", "IA", "MO")) == TRUE ~ "West North Central",
      (state %in% c("WI", "IL", "IN", "MI", "OH")) == TRUE ~ "East North Central",
      (state %in% c("KY", "TN", "MS", "AL")) == TRUE ~ "East South Central",
      (state %in% c("DE", "DC", "FL", "GA", "SC", "NC", "WV", "VA", "MD")) == TRUE ~ "South Atlantic",
      (state %in% c("CT", "ME", "MA", "VT", "NH", "RI")) == TRUE ~ "New England",
      (state %in% c("NJ", "NY", "PA")) == TRUE ~ "Middle Atlantic"
      
    )
  )
```

```{r}
holc_grades_race_region <- metro_grades_w_region %>% 
  group_by(holc_grade, region) %>% 
  summarise(
    pct_white = mean(pct_white),
    pct_black = mean(pct_black),
    pct_hisp = mean(pct_hisp),
    pct_asian = mean(pct_asian),
    pct_other = mean(pct_other))
```


## Mean racial composition percentages for A, B, C, and D HOLC Grades by State:
This code looks at the racial breakdown of HOLC grades, grouped by state. 

When arranging the results by HOLC A grades with the highest average white population percentage, we found that South Carolina is at the top with a figure of 88.98% — this is roughly 15 points higher than the national average for HOLC A grades, which is 73.78%. Other states that followed were Kentucky, Tennessee, Indiana and Virginia, which had white percentages 10 points greater than the national average. 

Conversely, states with Holc D grades that had the highest average Black population percentage include Mississippi, Arkansas, Alabama, Louisiana and North Carolina. Each state was at least 35 percentage points higher than the national average of the Black populations with a Holc D grade.

In our regional analysis, we previously found that Alabama and Mississippi have a less than 50% average white population but the region they're part of had the highest average percent of white population with A-level HOLC grades. The average Black populations in Alabama and Mississippi are 40% and 68%, respectively.

A scatterplot of the average Black population in all the states and the average Black population percent of HOLC D grades shows a strong positive correlation. A correlation test also shows a very low p-value, indicating the correlation between these two variables is statistically significant and probably not random.

Elsewhere, Maryland had the highest average Black percentage for HOLC A grades, with a figure of 41.04%. This is 35 points greater than the national average of 8.89%. Maryland's Black population is about 44%, so these numbers match pretty well.

```{r}

holc_grades_race_state <- metro_grades_plus_state %>% 
  group_by(holc_grade, state) %>% 
  summarise(
    avg_pct_white = mean(pct_white),
    avg_pct_black = mean(pct_black),
    avg_pct_hisp = mean(pct_hisp),
    avg_pct_asian = mean(pct_asian),
    avg_pct_other = mean(pct_other))

```

```{r}
metro_grades_all <- metro_grades_plus_state %>% 
  group_by(state) %>% 
  summarise(
    white_pct_pop = mean(pct_white),
    black_pct_pop = mean(pct_black),
    hisp_pct_pop = mean(pct_hisp),
    asian_pct_pop = mean(pct_asian),
    other_pct_pop = mean(pct_other)
  )
```


```{r}
compare_pop_and_grades <- holc_grades_race_state %>% left_join(metro_grades_all, by=c("state"="state"))
```

```{r}
compare_pop_and_grades_just_d <- compare_pop_and_grades %>% 
  filter(holc_grade == "D")
```


```{r}
compare_pop_and_grades_just_d %>%
  ggplot() +
  geom_point(aes(x=black_pct_pop,y=avg_pct_black)) +
  geom_smooth(aes(x=black_pct_pop,y=avg_pct_black), method="lm")
```

```{r}
cor.test(compare_pop_and_grades_just_d$black_pct_pop, compare_pop_and_grades_just_d$avg_pct_black)
```


#Explaination of outliers when compared to state's census demographics

Following an investigation that strictly analyzed HOLC neighborhood demographics, we joined our data with Census demographic data and compared racial compositions in HOLC A, B, C and D neighborhoods to the state-wide figures. 

Doing so provided interesting background for our findings in our previous sections. South Carolina, the outlier with an especially high percentage white population in its HOLC A neighborhoods, has a demographic composition not all that different from Maryland, the outlier with an especially high percentage Black population in its HOLC A neighborhoods. According to Census data, Maryland is 55.54% white and 29.89% Black, while South Carolina is 67.16% white and 26.77% Black. Now, although the white population percentages differ more substantially, the Black population percentages are only off by ~3 percentage points. Yet, their HOLC A grade compositions differ heavily from one another.

In an earlier section, we identified that the  West North Central region has an average white population percentage of HOLC D grades of 58.2% — a major outlier in the dataset. Comparing the states that comprise that region's Census data to their HOLC data helped explain why that is. Each state has a white population percentage greater than 80%. 

Summary statistics on this dataset also brought back its own unique findings. 23 of our 37 states with HOLC data had HOLC A demographics less white than their state's Census demographics. Elswhere, nine of 37 states had HOLC A demographics that were more Black than their state's Census demographics, including a chunk of West North Central states like South Dakota, Nebraska, and Iowa. Additionally, eight states' HOLC A demographics were more Hispanic than their state's demographics. This is likely explained by the increased diversity in metro-areas compared to the larger demographics of the state (https://www.usnews.com/news/cities/articles/2020-01-22/americas-cities-are-becoming-more-diverse-new-analysis-shows).


```{r}
us_pop_and_holc%>%
  filter(holc_grade=="A")%>%
  filter(total_pct_white>HOLC_pct_white)

us_pop_and_holc%>%
  filter(holc_grade=="A")%>%
  filter(total_pct_black<HOLC_pct_black)

us_pop_and_holc%>%
  filter(holc_grade=="A")%>%
  filter(total_pct_hispanic<HOLC_pct_hispanic)
```

```{r}
census_api_key("b83f426335eaf420358cb85a8df817d0f82924c5")
```
#Map displaying demographic difference between HOLC A grade and top four state-wide race demographics.
Through these Datawrapper maps, we visually display the differences between each state's HOLC A neighborhood composition and the state's Census demographics. To do so, we break down HOLC A neighborhoods by Black, white, Asian and Hispanic composition then represent the difference with the state's total racial composition by percentage points. These were the main takeaways:

1. Maryland and Midwest states have a higher percentage Black population in HOLC A neighborhoods than their respective state demographics. This means that neighborhoods that were considered desirable by the Home Owners Loan Corporation are now more diverse than their state average, an interesting finding, considering diverse neighborhoods were given lower grades during this New Deal-era program.

2. Texas and South Dakota HOLC A neighborhoods are much less white than state demographics. Such findings reveal, similar to the previous finding, that neighborhoods then considered desirable are now less white than the state average. Conversely, Southeastern states have whiter HOLC A Neighborhoods than their state demographics, possibly indicating that the status quo of HOLC grades and the New Deal-era remain.

3. California, Illinois, and New York HOLC A neighborhoods are all significantly less Hispanic than their state demographics. This is an interesting finding considering the overall diversity of these states. What this could indicate, however, is that these states' Hispanic populations are more concentrated outside of these metro areas, or that they are concentrated in HOLC B, C, and D neighborhoods.

4.Lastly, our maps show that Minnesota and New Hampshire lead states with HOLC A neighborhoods more Asian than their state demographics. 

Black HOLC A Grade Demographics vs. State Demographics
https://datawrapper.dwcdn.net/iQyV2/1/

White HOLC A Grade Demographics vs. State Demographics:
https://datawrapper.dwcdn.net/K4tQ0/3/

Hispanic HOLC A Grade Demographics vs. State Demographics:
https://datawrapper.dwcdn.net/73fPV/2/

Asian HOLC A Grade Demographics vs. State Demographics:
https://datawrapper.dwcdn.net/xEauK/1/

```{r}
v20 <- load_variables(2020, "acs5", cache = TRUE)
```

```{r}

#this code loads population totals for the four most populous races in our HOLC data
us_state_pop <- get_acs(geography="state", variables="B01003_001")%>%
  rename(total_pop=estimate)

us_state_pop_white <- get_acs(geography="state", variables="B01001A_001")%>%
  rename(white_pop=estimate)

us_state_pop_Black <- get_acs(geography="state", variables="B01001B_001")%>%
  rename(Black_pop=estimate)

us_state_pop_Hispanic <- get_acs(geography="state", variables="B01001I_001")%>%
  rename(Hispanic_pop=estimate)

us_state_pop_Asian <- get_acs(geography="state", variables="B01001D_001")%>%
  rename(Asian_pop=estimate)


#this code joins those population totals into one dataframe
us_state_pop <-us_state_pop%>%left_join(us_state_pop_white, by=c("NAME"))

us_state_pop <-us_state_pop%>%left_join(us_state_pop_Black, by=c("NAME"))

us_state_pop <-us_state_pop%>%left_join(us_state_pop_Hispanic, by=c("NAME"))

us_state_pop <-us_state_pop%>%left_join(us_state_pop_Asian, by=c("NAME"))

#this code helps load state abbreviations, trim data
us_state_pop<-us_state_pop%>%
  select(GEOID, NAME, total_pop, white_pop, Black_pop, Hispanic_pop, Asian_pop)%>%
  rename(state_name=NAME)

fips_codes<-fips_codes%>%
  select(state,state_name)%>%
  distinct()

us_state_pop <-us_state_pop%>%left_join(fips_codes, by=c("state_name"))

## I got an error here -Devon. ##Fixed it - Emmett
us_state_pop<-us_state_pop%>%
select(state, total_pop, white_pop, Black_pop, Hispanic_pop, Asian_pop)

#this code calculate percentage of total population
us_state_pop<-us_state_pop%>%
  mutate(pct_white=(white_pop/total_pop)*100)%>%
    mutate(pct_Black=(Black_pop/total_pop)*100)%>%
    mutate(pct_Hispanic=(Hispanic_pop/total_pop)*100)%>%
    mutate(pct_Asian=(Asian_pop/total_pop)*100)

us_state_pop<-us_state_pop%>%
  select(state,pct_white,pct_Black,pct_Hispanic,pct_Asian)

us_state_pop<-us_state_pop%>%
  rename(total_pct_white=pct_white)%>%
  rename(total_pct_black=pct_Black)%>%
  rename(total_pct_hispanic=pct_Hispanic)%>%
  rename(total_pct_asian=pct_Asian)


#this code joins our HOLC data with our census population data
us_pop_and_holc<-us_state_pop%>%left_join(holc_grades_race_state, by=c("state"))
  
us_pop_and_holc<-us_pop_and_holc%>%
  rename(HOLC_pct_white=pct_white)%>%
  rename(HOLC_pct_black=pct_black)%>%
  rename(HOLC_pct_hispanic=pct_hisp)%>%
  rename(HOLC_pct_asian=pct_asian)

  us_pop_and_holc<-us_pop_and_holc%>%
    rename(HOLC_pct_other=pct_other)
  
  #this code creates dataframe for each of four race columns ad calculates difference between percent of total population and percent of HOLC A grade population, by state
  us_white_holc_dif<-us_pop_and_holc%>%
    select(state,holc_grade,HOLC_pct_white,total_pct_white)%>%
    mutate(diff_btwn_total_pct_and_holc_pct = (total_pct_white - HOLC_pct_white))
  
 us_white_holc_dif <-us_white_holc_dif%>%
  filter(holc_grade=="A")

 
  us_black_holc_dif<-us_pop_and_holc%>%
    select(state,holc_grade,HOLC_pct_black,total_pct_black)%>%
    mutate(diff_btwn_total_pct_and_holc_pct = (total_pct_black - HOLC_pct_black))%>%
    filter(holc_grade=="A")
  
  us_hisp_holc_dif<-us_pop_and_holc%>%
    select(state,holc_grade,HOLC_pct_hispanic,total_pct_hispanic)%>%
    mutate(diff_btwn_total_pct_and_holc_pct = (total_pct_hispanic - HOLC_pct_hispanic))%>%
    filter(holc_grade=="A")
  
    us_asian_holc_dif<-us_pop_and_holc%>%
    select(state,holc_grade,HOLC_pct_asian,total_pct_asian)%>%
    mutate(diff_btwn_total_pct_and_holc_pct = (total_pct_asian - HOLC_pct_asian))%>%
    filter(holc_grade=="A")
    
write_csv(us_white_holc_dif, "us_white_holc_dif.csv")

write_csv(us_black_holc_dif, "us_black_holc_dif.csv")

write_csv(us_asian_holc_dif, "us_asian_holc_dif.csv")

write_csv(us_hisp_holc_dif, "us_hisp_holc_dif.csv")

```

###  What is the Holc grade breakdown by state?
This process made us realize that our dataset aggregates the A, B, C, and D HOLC grades for each city then summarizes the demographic data for each grade. Thus, when we went to summarize HOLC grades by state, we found that the grade counts were consistent across all columns. Our discovery leads us to a new line of inquiry, however. Next week, we can analyze each state's data in a way similar to the process in Sentence 1 and summarize the mean demographic percentages for each state's A, B, C, and D grades.
```{r}
# Put code to reverse engineer sentence here


holc_by_state <- metro_grades_plus_state %>%
  group_by(state) %>% 
  summarise(
    a_count = sum(holc_grade == "A"),
    b_count = sum(holc_grade == "B"),
    c_count = sum(holc_grade == "C"),
    d_count = sum(holc_grade == "D")
  )
```  

## What is the racial brakdown of each HOLC grade? 
We first tried to put count each row into these "buckets" based on the percent white population, and we were going to do something similar for each race. The code was not counting correctly though, and we couldn't figure out why. 
```{r}
#lets do quartiles of percentages !!!, also consider plurality, not just majority
#could do something like this but idk why it's not working. may just have to do four tables (filtering on each quartile) for each race and join them together to get what we want. i just feel like there's a more efficient way to do that - Devon

holc_count_white_neighborhoods <- metro_grades %>% 
  mutate(
  white_neighborhoods_count = case_when(
    (pct_white %in% c(0.00:25.00)) == TRUE ~ "0-25% white neighborhood",
    (pct_white %in% c(25.01:50.00)) == TRUE ~ "25-50% white neighborhood",
    (pct_white %in% c(50.01:75.00)) == TRUE ~ "50-75% white neighborhood",
    (pct_white %in% c(75.01:100.00)) == TRUE ~ "75-100% white neighborhood"))
 # group_by(white_neighborhoods_count) %>% 
  #summarise(
   # count = n()
  #)
```

We then tried to do the counting individually and join the tables together, but it was too much confusing code. 
```{r}
#holc grades with pct >20 per race

holc_count_black <- metro_grades %>% 
  group_by(holc_grade) %>% 
  filter(pct_black > 20) %>% 
  summarise(
    twenty_pct_black_neighborhoods_count = n()
  )

holc_count_hispanic  <- metro_grades %>% 
  group_by(holc_grade) %>% 
  filter(pct_hisp > 20) %>% 
  summarise(
    twenty_pct_hisp_neighborhoods_count = n()
  )


holc_count_white  <- metro_grades %>% 
  group_by(holc_grade) %>% 
  filter(pct_white > 20) %>% 
  summarise(
    twenty_pct_white_neighborhoods_count = n()
  )

holc_grades_twenty <- holc_count_white %>% inner_join(holc_count_black, by=c("holc_grade"))

holc_grades_twenty <-holc_grades_twenty%>%inner_join(holc_count_hispanic, by=c("holc_grade"))
```

#Conclusion

In summary, our investigation centered on the geographic differences of HOLC grades and the racial composition of HOLC neighborhoods. Our findings show that some states have more diverse HOLC A neighborhoods — those considered most desirable to live in — than their state demographics, which indicates diversifying neighborhoods in areas that were once judged negatively for diversity. Additionally, in Southeastern states like South Carolina, the high proportion of white residents in HOLC neighborhoods that once bolstered a neighborhood's A grade, are more white than their state population — a possible lasting side-effect from redlining. 


-30-
